Neural Networks for Neural Tumors

### 2025-06-01 latex/draft.tex
evaluation and conclusion draft

This paper by Khan et al. (2021) begins by discussing brain tumor classification CNN models and their accuracies, which range from 80\% to 98\%. The authors point out two key challenges when training such models: first, a lack of properly labeled data, which makes effective training a lot more expensive and imprecise; and second, inconsistencies in image sizes, since a requirement of CNNs is fixed size input. Their dataset is divided into three parts: training (for learning model weights), validation (for tuning model parameters), and testing (for final evaluation). The authors mentioned preparing the images using an open source computer vision tool called Canny Edge Detection. Due to the limited dataset size, data augmentation techniques such as flipping, rotation, and brightness adjustment are applied to increase data variability. The model uses the ReLU activation function and stochastic gradient descent for optimization.

Due to limited computing resources, our group will likely rely on a pre-trained model. The article mentioned eight layers of neurons so that is a good starting point for us. We have put some thought into data pre-processing and will look into Canny Edge Detection as mentioned in the article. The article also mentioned the concept of data augmentation, which we will certainly apply. Given our small dataset, applying these techniques could help improve generalization on unseen data. It was also nice to see familiar terms like ReLU and stochastic gradient descent. We will definitely give those a try and see how the techniques affect our model's performance.

@@

\section{\large Evaluation (1-2 pages)}
Include all 3 models and their confusion matrix results, do not include the actual confusion matrix, only the f1 scores, precision, etc.

\section{\large Conclusion (1/2 page)}
Talk about how we chose CNN as our final model and reiterate how it performs

>>>>

Khan et al. (2021) examine convolutional neural network (CNN) models for brain tumor classification, reporting classification accuracies ranging from 80\% to 98\%. The authors highlight two primary challenges in training such models: (1) the scarcity of accurately labeled data, which increases training complexity and cost, and (2) inconsistency in image dimensions, given that CNNs require fixed-size input. Their dataset is partitioned into training, validation, and testing sets, serving model training, hyperparameter tuning, and final evaluation, respectively. Image pre-processing is performed using Canny Edge Detection, an open-source computer vision technique. To address the limited dataset size, the authors employ data augmentation methods including flipping, rotation, and brightness adjustments to enhance variability. The model utilizes the ReLU activation function and stochastic gradient descent (SGD) for optimization.

Given our limited computational resources, we plan to utilize a pre-trained model. The eight-layer architecture discussed in the paper provides a useful baseline. We are also considering Canny Edge Detection for pre-processing, as suggested by the authors. Furthermore, we intend to implement data augmentation techniques to improve generalization on our small dataset. The use of ReLU and SGD aligns well with our current understanding, and we aim to assess their impact on our model’s performance.

@@

\section{\large Evaluation (1-2 pages)}
In this project, we used three learning models (CNN, SVM, and RF) and compared the results of each. The dataset comprises four categories: glioma, meningioma, pituitary, and no tumor.

The CNN model demonstrated the highest overall accuracy at 96.68\%, followed closely by SVM at 95.66\%, and RF at 94.90\%. Macro and weighted F1-scores mirrored this trend, with CNN achieving 0.97, SVM 0.96, and RF 0.95 in both averages. These scores indicate that CNN not only achieves better overall classification performance but also maintains consistency across classes in terms of precision and recall.

Diving into per-class specifics, we observe that for the label "glioma", CNN exhibited the highest precision (0.98) and recall (0.93), indicating strong identification capability with few false positives and a relatively low false negative rate. SVM matched CNN in precision (0.98) but had a lower recall (0.91), while RF also had a precision (0.98) but the lowest score in recall at 0.89. This suggests that both SVM and RF are more likely to miss "glioma" cases compared to CNN while matching its precision.

In the case of "meningioma", while CNN had the best precision (0.93), SVM and RF had equal or superior recall (0.97 and 0.98 respectively) compared to CNN (0.97). SVM and RF had reduced precision (0.89 and 0.87) which is a notable difference from CNN's aforementioned score. This pattern indicates that SVM and RF are more sensitive in detecting "meningioma" but may yield more false positives than CNN.

In classifying "no tumor", RF achieved perfect precision (1.00), meaning that every time RF predicted "no tumor", the model was correct. However, its recall (0.93) was the lowest among the three, pointing to a higher rate of false negatives. CNN balanced both precision (0.97) and recall (0.98) well, outperforming both alternatives in overall reliability for this class.

For the label "pituitary", all three models performed strongly, with CNN leading in both precision and recall (0.99 each). SVM and RF followed closely with both metrics at 0.98 and 0.97-0.98, respectively, indicating that all models are highly effective in classifying "pituitary" tumors.

\section{\large Conclusion (1/2 page)}
The evaluation highlight CNN's strength in handling complex, high-dimensional image data, likely due to its hierarchical feature extraction capability. Its superior balance in both precision and recall across all tumor types supports its suitability for real-world clinical deployment where both false positives and false negatives carry significant risk.

SVM performed comparably well, especially in the detection of meningioma and pituitary tumors, and may offer a lighter-weight alternative in environments with limited computational resources. However, it showed reduced precision in some classes, making it slightly less robust overall.

RF, while achieving remarkable precision, particularly for the "no tumor" class, exhibited reduced recall across multiple classes. This suggests a tendency to make fewer false positives but at the expense of missing actual tumor cases, which is a very critical and life-threatening issue in medical diagnostics.
